NMsim provides an interface to running modified Nonmem models. With Nonmem's `$DESIGN` features, simple NMsim calls can solve tasks such as 

- ...
- ...


`NMsim` can be used for other types of model executions than simulations. To run an estimation model on a provided data set, use `method.sim=NMsim_asis`:

```{r,eval=FALSE}
NMsim(file.mod,
      data,
      method.sim=NMsim_asis)
```

If you are on linux/unix and have access to a sge-type cluster, you
can add `sge=TRUE` to submit to the cluster using `qsub`. 

### Generate data set

### Modify sections
The `modify.model` 
To rewrite the `$ESTIMATION` section to use FOCEI, you can simply overwrite the `$ESTIMATION` section:


```{r,eval=FALSE}
NMsim(file.mod,
      data,
      method.sim=NMsim_asis,
      sge=TRUE,
      modify.model=list(ESTIMATION="$ESTIMATION METHOD=COND INTER MAXEVAL=9999 NOABORT NSIG=2 SIGL=6 PRINT=5 POSTHOC")
      )
```

The `$DESIGN` section can similarly be inserted.

### Modify and fix parameter values
The coming NMsim 0.1.6 and accompanying NMdata 0.1.9 provide a simple
interface to controlling the parameter sections `$THETA`, `$OMEGA` and
`$SIGMA`. The default is to update these sections with the estimated
values in a `.ext` file which NMsim will look for next to the
simulated control stream, if not instructed otherwise. To fix
`THETA(1)` to the estimated value, you can do:


```{r,eval=FALSE}
NMsim(file.mod,
      data,
      method.sim=NMsim_asis,
      sge=TRUE,
      inits=list(method="nmsim2","theta(2)=list(fix=1)")
      )
```
