---
title: "Simulation-based forest plots with NMsim"
author: 
  - name: "Philip Delff"
  - name: "Boris Grinshpun"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
toctitle: "Contents"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: false
    df_print: paged
editor_options:
  chunk_output_type: inline
  code_folding: show
  bibliography: [packages.bib]
pkgdown:
  as_is: true
---

# Introduction
The forest plot is an effective and widely recognized way to
illustrate estimated covariate effects on exposure or response, or parameters
related to these (e.g. clearance). The forest plot typically includes precision of the estimate in terms of a confidence interval. Often, an acceptance region such as the 80%-125% bio equivalence is included for comparison.

In some cases, the forest plot can be derived based on model estimates
without simulation. This is the case for a forest plot on exposure if
the PK is linear and only steady-state average concentration (or AUC)
is of interest. If the PK is non-linear and/or exposure metrics such
as Cmax which depends on multiple PK parameters are of interest, a
simulation-based forest plot may be needed. As we shall see, NMsim
provides a flexible and concise framework to perform the required
simulations, a function for summarizing the simulation results is
provided, and it is demonstrated how the summary is easily plotted
using the coveffectsplot R package.

This vignette is no attempt to harmonize how to construct a forest
plot. Instead, we shall describe the steps taken, and it is up to the
scientist to choose how they want to construct the analysis to address
their questions most effectively. 

# Initialization

```{r,setup,echo=TRUE}
library(data.table)
library(NMsim)
library(NMdata)  
library(NMcalc)  ## Optional. Used to calculate AUC.
library(coveffectsplot)

NMdataConf(
    path.nonmem = "/opt/NONMEM/nm75/run/nmfe75", ## path to NONMEM executable
    dir.sims="simtmp-forest", ## where to store temporary simulation results
    dir.res="simres-forest",   ## where to store final simulation results
    as.fun="data.table"
)

```

A model is selected. 
```{r}
file.mod <- "NMsim-forest-models/xgxr134.mod"
```


# Generation of simulation input data

The simulation data set has to match the model in compartment numbers, and it must contain all variables needed to run the NONMEM model. We simulate daily dosing of 30 mg. 



```{r}

doses <- NMcreateDoses(TIME=0,AMT=30,addl=list(ADDL=100,II=24),col.id=NA)

```

`NMsim::expandCovs()` can be used to construct a set of simulations needed to derive a forest plot. It varies one covariate at a time, keeping all other covariates at their reference value. It can derive references and quantiles from a data set.

For reference values we use median in the observed population for continuous covariates and manually select "Female" as the reference for sex.


```{r}
## reading output and input tables from estimation. Used to determine
## reference values and quantiles.
data.ref <- NMscanData(file.mod)
covs <- expandCovs(
    AGE=list(ref=median,quantiles=c(10,25,75,90)/100,label="Age (years)"),
    ## notice, values OR quantiles can be provided
    WEIGHTB=list(ref=median, quantiles=c(10,25,75,90)/100, label="Bodyweigt (kg)"),
    MALEN=list(ref=c(Female=0), values=c(Male=1), label="Sex"),
    data=data.ref,
    as.fun="data.table"
)

covs[,ID:=.GRP,by=.(type,covvar,covval)]


dt.dos <- covs[,doses[],by=covs]

```

We sample every 15 minutes for one day in steady-state.

```{r}
## add a sampling scheme
time.sim <- data.table(TIME=seq(0,24,.25)+30*24,period="Steady-State")
dt.sim <- addEVID2(dt.dos,TIME=time.sim,CMT=2)
```

```{r,eval=FALSE}
ext <- sampleParsSimpar(file.mod=file.mod,nsim=500,seed.R=6789)

## library(devtools)
## load_all("~/wdirs/NMsim")

if(F){
    simres.forest <- NMsim(file.mod # path to NONMEM model
                          ,data=dt.sim, # simulation dataset
                          ,name.sim="forest_simpar" # output name suffix
                           ##,method.sim=NMsim_VarCov # sampling method
                          ,method.sim=NMsim_VarCov # sampling method
                          ,ext=ext
                          ,typical=TRUE # FALSE to include BSV
                          ,table.vars=cc(PRED,IPRED) # output table variables
                          ,method.update.inits="nmsim" # update parameters from ext
                           ## ,nsims=2 # number of simulations
                           ## ,subproblems=10 # number of simulations
                          ,seed.R=342 # seed for reproducibility
                          ,sge=TRUE # TRUE if submitting to a cluster
                          ,nc=1
                           ## ,reuse.results=TRUE # TRUE to load existing results
                          ,quiet = FALSE ## FALSE to view all output messages
                          ,reuse.results=TRUE
                           )
}

simres <- NMreadSim(simres.forest,wait=T,rm.tmp=TRUE)

```

```{r}
simres <- NMreadSim("simres-forest/xgxr134_forest_simpar_MetaData.rds")
## source("~/wdirs/NMpub/ACOP2024/forest/forest_summarize.R")

########## postprocess
### Exposure metrics
funs.exposure <- list(
    "Cmax"=function(x) max(x$PRED)
   ,"AUC"=function(x) trapez(x$TIME,x$PRED)
    ## ,"Concentration at 4 hours"=function(x) x$value[x$TAPD==4]
)



sum.uncertain <- NMsim:::summarizeCovvar(simres,
                                         funs.exposure = funs.exposure,
                                         by=cc(period),
                                         cover.ci=.95
                                         )

```

Plotting using `coveffectsplot`.

```{r}

sum.uncertain.plot <- copy(sum.uncertain)

setnames(sum.uncertain.plot,
         cc(covvalf,predmm,predml,predmu,metric.var),
         cc(label,mid,lower,upper,paramname)
         )

sum.uncertain.plot[,MEANVAL:=mid]
nsig <- 3
sum.uncertain.plot[,LABEL := sprintf("%s [%s - %s]",signif2(mid,nsig),signif2(lower,nsig),signif2(upper,nsig))]
sum.uncertain.plot[,covname:=covlabel]



escape_newline <- function(s) {
    gsub("\\\\n", "\\\n", s)
}

### where is this supposed to come from?
label_value <- function(x,... )x

textsize <- 10
forest1 <- forest_plot(
    data = sum.uncertain.plot,
    facet_formula = "covlabel ~ paramname", 
    facet_scales = "free_y",
    facet_space = "free_y",
    ## facet_labeller = "label_value",
    xy_facet_text_bold = FALSE, 
    plot_table_ratio = 1.7, 
    table_text_size = 3,
    ## label_wrap_width = 16L,
    x_label_text_size = textsize,
    y_label_text_size = textsize, 
    ## table_title_size = 16L,
    x_facet_text_size = textsize, 
    y_facet_text_size = textsize,
    base_size=textsize,
    ## interval_size=.65, ## geom_pointerinterval size
    strip_placement = "outside",
    table_position = "right",
    legend_order = c("pointinterval", "ref", "area"),
    x_range = c(.5,1.5),
    ## interval_size=1.1,
    ## interval_fatten = 5L,
    ## interval_linewidth = 2,
    ref_legend_text = escape_newline("Reference (vertical line)\\nClinically relevant limits 0.8-1.25 (colored area)"),
    area_legend_text = escape_newline("Reference (vertical line)\\nClinically relevant limits 0.8-1.25 (colored area)"),
    facet_switch = c("y"),
    legend_position="bottom"
)


## forest1


```
