---
title: "Simulate Known Subjects Using Emperical Bayes Estimates (Etas/Phis)"
author: 
  - name: "Philip Delff"
    date: '`r format(Sys.Date(), "%B %d, %Y")`'
    toctitle: "Contents"
    output:
      html_document:
        toc: true
        toc_depth: 4
        toc_float: true
        number_sections: false
        df_print: kable
        editor_options:
          chunk_output_type: inline
          code_folding: show
          bibliography: [packages.bib]
          pkgdown:
            as_is: true
---

```{r,include = FALSE}
##knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(
                      collapse = TRUE
                     ,comment = "#>"
                     ,fig.width=7
                     ,cache=FALSE
                  )

## this changes data.table syntax. I think we can do without.
## knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r,setup,include=F}
## library(devtools)
## unloadNamespace("NMsim")
## unloadNamespace("NMdata")

## load_all("~/wdirs/NMdata")
## load_all()
library(NMsim)
library(data.table)
library(NMdata)
## library(dplyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(tracee)
## library(tidyr)
library(fst)
library(knitr)
paths.nm <- c("/opt/NONMEM/nm75/run/nmfe75",
              "/opt/nonmem/nm751/run/nmfe75")
found.nm <- which(file.exists(paths.nm))
if(length(found.nm)){
    path.nonmem <- paths.nm[min(found.nm)]
} else {
    path.nonmem <- NULL
}

NMdataConf(
    path.nonmem = path.nonmem, ## path to NONMEM executable
    dir.sims="simtmp-known", ## where to store temporary simulation results
    dir.res="simres-known"   ## where to store final simulation results
)
theme_set(theme_bw())
this.script <- "NMsim-methods.Rmd"
writeOutput <- TRUE
file.project <- function(...)file.path(system.file("examples",package="NMsim"),...)
## file.project <- function(...)file.path("../inst/examples",...)
## file.project <- function(...)file.path("~/wdirs/NMsim/inst/examples",...)

found.files <- list.files(file.project("nonmem/NMsim"),pattern="noname\\.(lst|xml|ext|cov|cor|coi|phi|msf|msfi|msfo|tab)",full.names=TRUE)
unlink(found.files)

run.simuls <- FALSE
NMdataConf(as.fun="data.table")

dat.sim <- read.csv(system.file("examples/derived/dat_sim1.csv",package="NMsim"))
```


Built `r Sys.Date()` using NMsim `r packageVersion("NMsim")`.


## Objectives
This vignettes aims at enabling you to use `NMsim` for the following purposes

* Simualation of known subjects (estimated random effects), 


## Prerequisites
Simulation with known ETA values is only possible through NMsim's internal Nonmem execution method. That means, we need to provide the path to the Nonmem executable.

```{r,eval=TRUE}
NMdataConf(path.nonmem="/opt/NONMEM/nm75/run/nmfe75")
```
More information on configuration and test of configuration at
[`NMsim-config.html`](https://nmautoverse.github.io/NMsim/articles/NMsim-config.html). 

You should be familiar with basic `NMsim` arguments as described in
[`NMsim-intro.html`](https://nmautoverse.github.io/NMsim/articles/NMsim-intro.html). In
that vignette you should have learned to use the default simulation
method. While the default simulation method samples new subjects (using the `$OMEGA` metrix). This vignette provides various examples on how the `NMsim_EBE` simulation method can be used to simulate known subjects.


## Simulation of known subjects 
We sometimes want to simulate the observed subjects, i.e. the subjects we included in the estimation data. We want to reuse the estimated random effects (ETA's) given the
subject ID's. `NMsim` has a method for this called `NMsim_EBE`. The restriction is that
all subjects (values of `ID`) in the simulation input data must have
been used in the estimation input data.

For
`NMsim_EBE` to work, `ID` values in the simulation data set must be
identical to `ID` values used in the estimation.  For
models estimated with FOCEI, the ETAs are stored in the
`.phi` file produced by Nonmem. This file must be available, and NMsim will automatically use it if applicable.

For SAEM/IMP-based estimation NMsim cannot use the `.phi` file, and the user must have stored the ETAs in output
table files for the method to work. It does not matter where in the
table files they are, they just all need to be there. You can write
all of them like it is done in the example model provided with `NMsim`
called `xgxr032.mod`.

If using SAEM/IMP, remember to explicitly write ETAs to an output table in the Nonmem control stream:
```
$TABLE ETAS(1:LAST) NOAPPEND NOPRINT FILE=xgxr032_etas.txt
```

So a "subject" essentially means a set of ETAs. Covariates must still
be provided in the input data set and can be modified if
wanted. Notice, `NMsim_EBE` does nothing to restore covariates - you
must define covariates in input data as needed.


## Simulate known subjects on a new dosing regimen and new sample schedule
The following code takes an already created simulation data set with a
single ID, and merges all other columns than `ID` onto the observed
`ID`s. That gives the same simulation data for all subjects. 

First thing, we decide on a model (an input control stream of an estimated model) to use for the example:
```{r}
file.mod <- system.file("examples/nonmem/xgxr134.mod",package="NMsim")
```

We generate a simulation data set with a loading dose (300 mg) followed by 6 doses (150 mg) every 24 hours. For smoothness plotted curves we use hourly sampling.

```{r,datacreate}
doses <- NMcreateDoses(TIME=c(0,24),AMT=c(300,150),ADDL=5,II=24,CMT=1)
## Now we add the sample records using `addEVID2()`. 
dat.sim <- NMaddSamples(doses,TIME=0:(24*7),CMT=2)
```

```{r,eval=FALSE}
## read model results just to extract the observed ID's. Notice, if
## your model uses covariates, those should be retrieved too.
res.mod <- NMscanData(file.mod,quiet=TRUE)
ids <- unique(res.mod[,c("ID","WEIGHTB","AGE","MALEN")])

## Repeat the simulation data set for each ID and order accordingly

## we will merge with all the observed subjects in stead of the single
## ID in the original data set.
dat.sim$ID <- NULL
dat.sim.known <- egdt(ids,
                      dat.sim
                      )
setorder(dat.sim.known,ID,TIME,EVID)
## check data
NMcheckData(dat.sim.known,type.data="sim")

simres.known <- NMsim(file.mod=file.mod,
                      data=dat.sim.known,
                      method.sim=NMsim_EBE,
                      table.vars="PRED IPRED CL V2 KA",
                      name.sim="known1"
                      )
```

```{r,include=FALSE,eval=TRUE}
simres.known <- NMreadSim("simres-known/xgxr134_known1_MetaData.rds")
```

And the simulation results are plotted for each subject.
```{r,eval=TRUE}
ggplot(subset(simres.known,EVID==2),aes(TIME,IPRED,colour=factor(ID)))+
    geom_line()+
    theme(legend.position="none")
```

<!-- Show a plot of the individual parameters in estimate and in sim -->

## Individual dosing history, sample times or both
Individual simulations are useful for several purposes. Examples

* Reuse individual dosing history and use a new common sample
scheme. This could be for homogeneous evaluation of exposure metrics
such as Cmax, AUC, or just to show concentration-time profiles at
identical sampling times based on a model estimated on data with
heterogeneos sampling. Currently, there is no example of this in
this vignette but see the other examples, and it should be pretty
clear how to do this.

* Reuse dosing history and use individual sampling scheme from a
different data set, e.g. PD.

* Simulate a new common dosing and sampling scheme to simulate already
observed subjects on a new regimen. This is sometimes done if for
one reason or the other the model is not considered reliable for
simulation of new subjects but the individual parameter estimates
are trusted. 

* Reuse a simulated population. One may prefer to reuse the same
simulated subjects in multiple simulations for reproducibility and to have all difference
between say simulation results of different regimens be driven by
differences in the regimen, and not in the populations. This
use of `NMsim_EBE` is on the todo list tog get it's own vignette
[here](https://nmautoverse.github.io/NMsim/articles/NMsim-ReuseSimSubjects.html).


### Simulate individual dosing history at new individual sampling times for a PK/PD dataset
We want to plot some PD data angainst
PK. However, PD was sampled differnetly than PK, and we want to
evaluate the individual predictions of the PK model at the individual
PD samplng times - reusing the individual dosing history.

Reading some example PD data:
```{r,pddata,include=TRUE,eval=TRUE}
pd <- readRDS(system.file("examples/data/xgxr_pd.rds",package="NMsim"))
## some code below makes use of data.table features so we make it a data.table
setDT(pd)
```

For a PK model without time-varying covariates, suggested steps 
to generate the data for the simulation are:

* Take dose records from PK model estimation input data (`pkdos`). Just keep necessary columns like `ID`, `TIME`, `EVID`, `CMT`, `AMT`, `ADDL`, `II`, and any necessary covariates
* Take PD data observation records (`pdsamples`). Just keep `ID`, `TIME`, and set `EVID=2`.
* Add a unique row identifier to `pdsamples` (an integer row counter, like `ROW=1:nrow(pdsamples)`)
* Stack (`rbind` for data.tables or `bind_rows` in tidyverse) `pkdos` and `pdsamples` to one data set (`pdsim`)
* In `pdsim`, set `DV=NA`
* Sort `pdsim` at least by `ID`, `TIME` and `EVID`. There could be more depending on trial design

In case of time-varying covariates, you can keep all data records from the PK data (without `DV`), but change observation records to simulation records (`EVID=2` instead of `EVID=0`).

```{r,known-pkpd,eval=TRUE}
## Take dose records from PK model estimation input data
pkres <- NMscanData(file.mod,quiet=TRUE)
pkdos <- pkres[EVID==1,.(ID, TIME, EVID, CMT, AMT)]
## Take PD data observation records (`pdsamples`)
pdsamples <- pd[EVID==0,.(ID,TIME,LIDV)]
## Stack `pkdos` and `pdsamples` to one data set (`pdsim`)
pdsim <- rbind(pkdos,pdsamples,fill=TRUE)
## Nonmem needs a DV column to run
pdsim[,DV:=NA]
## only include subjects that were included in the PK model
pdsim <- pdsim[ID%in%pkres$ID]
setorder(pdsim,ID,TIME,EVID)
```

Then run `NMsim` like this. We are renaming the prediction columns using Nonmem `$TABLE` statement not to confuse PK predictions and PD predictions.

```{r,known-pkpd-run,eval=FALSE}
simres.pksim <- NMsim(file.mod,
                      data=pdsim,
                      name.sim="pkpd",
                      method.sim=NMsim_EBE,
                      table.vars="PKIPRED=IPRED PKPRED=PRED"
                      )
```
```{r,include=FALSE,eval=TRUE}
file.fst <- "simulate-results/xgxr021_pkpd_paths.rds"
simres.pksim <- NMreadSim(file.fst)
```



```{r,eval=TRUE}
ggplot(simres.pksim[!is.na(LIDV)&!is.na(PKIPRED)],aes(PKIPRED,LIDV))+
    geom_point()+
    labs(x="Individual PK prediction",y="Observed PD value")
``` 


